A thermodynamic view of dusty protoplanetary disks: axisymmetric stability

### 2016-11-09 summary.tex
tiny edits

\section{Summary}\label{summary}
In this paper we describe the similarity between an isothermal 
dusty gas and an adiabatic, pure gas. The correspondence arises
because drag forces reduce the relative velocity between gas and
dust. In the limit of perfect dust-gas coupling, $\tstop \to 0$,  
 dust is entrained in 
the gas, and the dust-to-gas ratio $\rhod/\rhog$ is conserved
following the dusty flow. This is analogous to entropy being conserved
following a adiabatic, pure gas.

For finite dust-gas drag, $\tstop\neq 0$, the dust content of a 
parcel of the mixture no longer conserved. The parcel 
can exchange dust particles with neighboring parcels owing 
to dust-gas relative drift. This is analogous to heat exchange between
a pure gas parcel and its surroundings.

We explicitly show that for a locally isothermal, dusty gas that the 
evolutionary equation of  $\rhod/\rhog$ may be replaced by an 
effective energy equation. This leads to a 
natural definition of the effective entropy of isothermal dusty gas as  
\begin{align*}
  \seff  = \ln{\left(\frac{c_s^2\rhog}{\rhog+\rhod}\right)}.  
\end{align*}
%This allows us to define the buoyancy frequency of an isothermal 
%dusty gas. 
Finite dust-gas friction then appears as an energy
source term.  This analogy with standard 
hydrodynamics with cooling/heating allow us to obtain dusty analogs of gaseous
instabilities, and provide thermodynamical interpretations of  
dust-drag instabilities.

@@

We  apply our thermodynamic framework to interpret
the streaming instability \citep[SI, ][]{youdin05a,jacquet11} and to generalize the vertical shear
instability \citep[VSI, ][]{nelson13,lin15} to dusty disks. We
explicitly show that the SI occurs because the evolution of gas
pressure lags behind  dust density; a general property of \emph{all}
instabilities driven by strong dust-gas drag. 
It takes a finite time for the gas to respond to
the dust motion. A lag implies a time interval when
gas pressure is still increasing while dust is already being expelled. 
The dusty gas then does positive work, which goes into amplifying
oscillation amplitudes. This interpretation is analogous to stellar pulsational
instabilities \citep{cox67}.

For the VSI we find dust-loading is generally stabilizing. In our disk models 
dust-loading does not affect VSI growth rates significantly, but
meridional motions may be suppressed where the dust-induced
vertical buoyancy dominates over vertical shear, consistent with our
previous study \citep{lin15}.

>>>>

\section{Summary}\label{summary}
In this paper we develop an analogy between isothermal 
dusty-gas and adiabatic, pure gas. The correspondence arises
because drag forces reduce the relative velocity between gas and
dust. In the limit of perfect dust-gas coupling, $\tstop \to 0$,  
 dust is entrained in 
the gas. Then the dust-to-gas ratio $\rhod/\rhog$ is conserved
following the flow. This is analogous to entropy being conserved
following a adiabatic, pure gas.

For finite dust-gas drag, $\tstop\neq 0$, the dust content of a 
parcel of the mixture is no longer conserved. The parcel 
can exchange dust particles with neighboring parcels. % owing 
%to dust-gas relative drift. 
This is analogous to heat exchange between a parcel of pure gas and its surroundings.

We explicitly show that for a fixed gas equation of state, the  
evolutionary equation for $\rhod/\rhog$ may be replaced by an 
effective energy equation. This leads to a 
natural definition of the effective entropy of isothermal dusty gas as  
\begin{align*}
  \seff  = \ln{\left(\frac{c_s^2\rhog}{\rhog+\rhod}\right)}.  
\end{align*}
%This allows us to define the buoyancy frequency of an isothermal 
%dusty gas. 
Finite dust-gas friction then appears as an energy
source term.  This analogy with standard 
hydrodynamics with cooling/heating allow us to obtain dusty analogs of gaseous
instabilities, and provide thermodynamical interpretations of  
dust-drag instabilities.

@@

We  apply our thermodynamic framework to interpret
the streaming instability \citep[SI, ][]{youdin05a,jacquet11} and to generalize the vertical shear
instability \citep[VSI, ][]{nelson13,lin15} to dusty disks. We
explicitly show that the SI occurs because the evolution of gas
pressure lags behind  dust density; a general property of \emph{all}
instabilities driven by strong dust-gas drag. 
It takes a finite time for the gas to respond to
the dust motion. A lag implies there exists a time interval where 
the gas pressure of a parcel of the mixture is increasing whilst dust is already being expelled.  
The dusty gas then does positive work, which goes into amplifying
oscillation amplitudes. This interpretation is analogous to stellar pulsational
instabilities \citep{cox67}.

For the VSI we find dust-loading is generally stabilizing. In our disk models 
dust-loading does not affect VSI growth rates significantly, but
meridional motions may be suppressed where the dust-induced
vertical buoyancy dominates over vertical shear, consistent with our
previous study \citep{lin15}. Since the dust-induced buoyancy forces increase away from the midplane, we find dust-loading can stabilize `surface modes' of the VSI, that would otherwise have the largest growth rates.
